---
title: "Analysis Factorial y Regresion logistica - CENTRO"
author: "Viviana Carrizales Luna"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Funciones
```{r}
asignar_etapa <- function(df) {
  inicio_construccion <- as.Date("2022-08-30")
  inicio_construccion2 <- as.Date("2024-02-24")
  
  df <- df %>% mutate(
    date = as.Date(date),
    etapa = case_when(
      date < inicio_construccion ~ 0,
      date >= inicio_construccion & date < inicio_construccion2 ~ 1,
      date >= inicio_construccion2 ~ 2
    )
  )
  
  return(df)
}

pruebas_no_parametricas <- function(df, contaminantes, tipo = "independiente", etapa_col = "etapa") {
  resultados <- data.frame()
  
  etapas <- sort(unique(df[[etapa_col]]))
  
  for (c in contaminantes) {
    for (i in 1:(length(etapas)-1)) {
      for (j in (i+1):length(etapas)) {
        grupo1 <- na.omit(df[df[[etapa_col]] == etapas[i], c])
        grupo2 <- na.omit(df[df[[etapa_col]] == etapas[j], c])
        
        if (tipo == "independiente") {
          test <- wilcox.test(grupo1, grupo2, exact = FALSE)  # Mann-Whitney
        } else if (tipo == "pareado") {
          n <- min(length(grupo1), length(grupo2))
          test <- wilcox.test(grupo1[1:n], grupo2[1:n], paired = TRUE)
        } else {
          stop("tipo debe ser 'independiente' o 'pareado'")
        }
        
        resultados <- rbind(resultados, data.frame(
          Contaminante = c,
          Comparacion = paste(etapas[i], "vs", etapas[j]),
          Estadistico = test$statistic,
          p_valor = test$p.value
        ))
      }
    }
  }
  
  return(resultados)
}

pca_rotado <- function(df, variables, nfactors = 3, rotate = "varimax") {
  # Estandarizar variables
  df_scaled <- scale(df[variables])
  
  # rotación
  pca_res <- principal(df_scaled, nfactors = nfactors, rotate = rotate, scores = TRUE)
  
  # Cargas
  cargas <- as.data.frame(pca_res$loadings[, 1:nfactors])
  colnames(cargas) <- paste0("PC", 1:nfactors)
  rownames(cargas) <- variables
  
  # Scores
  scores_df <- as.data.frame(pca_res$scores)
  colnames(scores_df) <- paste0("PC", 1:nfactors)
  
  # Unir scores al df original
  df_out <- df
  df_out <- cbind(df_out, scores_df)
  
  return(list(cargas = cargas, df_out = df_out, pca_obj = pca_res))
}

segmentar_pc_por_etapa <- function(df, pcs = c("PC1", "PC2", "PC3"), etapa_col = "etapa") {
  df_seg <- df %>% select(all_of(c(pcs, etapa_col)))
  return(df_seg)
}

prueba_kruskal_por_etapa <- function(df, factor) {
  kruskal.test(as.formula(paste(factor, "~ etapa")), data = df)
}

```
main
```{r}
library(dplyr)
library(psych)
library(ggplot2)
library(FSA) 
library(readxl)

df <- read_excel("CENTRO_AVENIDAS.xlsx")
variables <- c("CO_24h", "PM10_24h", "PM2.5_24h", "SO2_24h", "NO2_24h")
df <- df %>% select(all_of(c(variables, "date")))

df <- asignar_etapa(df)

# PCA rotado
pca_res <- pca_rotado(df, variables, nfactors = 3, rotate = "varimax")
cargas <- pca_res$cargas
df_scores <- pca_res$df_out


df_seg <- segmentar_pc_por_etapa(df_scores, pcs = c("PC1", "PC2", "PC3"))

# Kruskal–Wallis por PC1 y PC2
res_pc1 <- prueba_kruskal_por_etapa(df_seg, factor = "PC1")
res_pc2 <- prueba_kruskal_por_etapa(df_seg, factor = "PC2")
res_pc3 <- prueba_kruskal_por_etapa(df_seg, factor = "PC3")

pca_res

```
# plot de las time series de contaminantes

```{r}
# install.packages(c("tidyverse", "viridis"))  # run once if needed
library(tidyverse)
library(viridis)   # for colorblind-friendly palette

# --- Ensure there's exactly one Date column named `date` ---
df2 <- if ("date" %in% names(df)) {
  df %>% mutate(date = as.Date(date))
} else {
  df %>% tibble::rownames_to_column("date") %>% mutate(date = as.Date(date))
}

# --- Drop data from Jun 2024 onward ---
df_long <- df2 %>%
  filter(date < as.Date("2024-06-01")) %>%
  pivot_longer(
    cols = -c(date, etapa),   # 👈 excluimos 'date' y 'etapa'
    names_to = "pollutant",
    values_to = "value"
  ) %>%
  mutate(
    pollutant = str_replace(pollutant, "_.*", "")  # 👈 reemplaza "SO_24h" -> "SO"
  )

# --- Plot: all pollutants in one elegant time series ---
p <- ggplot(df_long, aes(x = date, y = value, color = pollutant)) +
  geom_line(linewidth = 1.1, alpha = 0.95) +
  scale_color_viridis_d(name = "Pollutant", end = 0.9,
                        guide = guide_legend(title.position = "top")) +
  labs(
    x = "Date",
    y = "Concentration (µg/m³)"
  ) +
  scale_x_date(date_labels = "%b %Y") +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(face = "bold"),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

p

# Optional: save a high-res image for your report
ggsave("pollutants_timeseries.png", p, width = 12, height = 6, dpi = 300)


```
```{r}
R  <- cor(df[variables])
ev <- eigen(R, symmetric = TRUE, only.values = TRUE)$values
p  <- length(ev)
pve <- ev / sum(ev)            # % variance per PC
cum <- cumsum(pve)

# --- 2) Plot (ggplot2) ---
library(ggplot2)
library(scales)

top <- min(12, p)              # show first 12 (or fewer) components
d <- data.frame(
  PC     = factor(paste0("PC", seq_len(p)), levels = paste0("PC", seq_len(p))),
  PVE    = pve,
  CumPVE = cum
)

ggplot(d[1:top, ], aes(PC, PVE)) +
  geom_col(width = 0.7) +
  geom_point(aes(y = CumPVE), size = 2) +
  geom_line(aes(y = CumPVE, group = 1), linewidth = 0.6) +
  geom_text(aes(label = percent(PVE, accuracy = 0.1)),
            vjust = -0.6, size = 3.2) +
  geom_hline(yintercept = 1/p, linetype = "dashed", linewidth = 0.35) +
  annotate("text", x = top - 0.25, y = 1/p + 0.02,
           label = "Kaiser: λ=1 ⇒ 1/p",
           size = 3.2, hjust = 1) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0, 1.05),
                     expand = expansion(mult = c(0, .02))) +
  labs(title = "Scree Plot (Correlation Matrix PCA)",
       subtitle = "Bars: % variance by PC · Line: cumulative %",
       x = "Principal Component",
       y = "Proportion of Variance Explained") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    axis.title.x = element_text(margin = margin(t = 6)),
    axis.title.y = element_text(margin = margin(r = 6))
  )

# --- 3) Save for your report (high-res) ---
ggsave("scree_corr_matrix_centro.png", width = 8, height = 5, dpi = 300)
```
```


```{r}
res_pc3
```
```{r}
df_seg$etapa <- as.factor(df_seg$etapa)
dunnTest(PC1 ~ etapa, data = df_seg, method = "bonferroni")
```

```{r}
df_seg$etapa <- as.factor(df_seg$etapa)
dunnTest(PC2 ~ etapa, data = df_seg, method = "bonferroni")
```

```{r}
df_seg$etapa <- as.factor(df_seg$etapa)
dunnTest(PC3 ~ etapa, data = df_seg, method = "bonferroni")
```


```{r}
cargas_df <- as.data.frame(cargas)
cargas_df$Variable <- rownames(cargas_df)

cargas_long <- cargas_df %>%
  pivot_longer(cols = starts_with("PC"),
               names_to = "Component", values_to = "Loading") %>%
  mutate(Component = recode(Component,
                            "PC1" = "Factor 1",
                            "PC2" = "Factor 2",
                            "PC3" = "Factor 3"))

p_loadings <- ggplot(cargas_long, aes(x = Variable, y = Loading, fill = Component)) +
  geom_col(position = "dodge") +
  labs(title = "Aporte de los contaminantes para cada factor",
       x = "Variables", y = "Loading") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("pca_loadings_centro.png", p_loadings, width = 10, height = 6, dpi = 300, bg = "transparent")

```

```{r}
library(tidyverse)

# Reshape, rename, and ensure etapa is categorical
df_long <- df_scores %>%
  mutate(etapa = as.factor(etapa)) %>%   # <-- ensure etapa is factor
  pivot_longer(cols = starts_with("PC"),
               names_to = "Factor",
               values_to = "Score") %>%
  mutate(Factor = recode(Factor,
                         "PC1" = "Factor 1",
                         "PC2" = "Factor 2",
                         "PC3" = "Factor 3"))

# Boxplots by etapa
p_box <- ggplot(df_long, aes(x = etapa, y = Score, fill = etapa)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.size = 2) +
  facet_wrap(~Factor, scales = "free_y") +
  scale_fill_brewer(palette = "Set2", name = "Etapa") +
  labs(
    title = "Distribution of Factor Scores by Etapa",
    x = "Etapa",
    y = "Score"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.title   = element_text(face = "bold"),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

print(p_box)

# Save with transparent background
ggsave("factor_boxplots_centro.png", p_box, width = 10, height = 6, dpi = 300, bg = "transparent")


```

```{r}
res_pc1
```
```{r}
res_pc2
```

```{r}

library(dunn.test)
# Dunn post-hoc para PC1 agrupado por Etapa
dunn_result <- dunn.test(df_seg$PC1, g = df_seg$etapa, method = "bonferroni")

# Mostrar resultados
dunn_result
```

```{r}

library(dunn.test)
# Dunn post-hoc para PC1 agrupado por Etapa
dunn_result <- dunn.test(df_seg$PC2, g = df_seg$etapa, method = "bonferroni")

# Mostrar resultados
dunn_result
```
## grafico de intervalos
```{r}

library(dplyr)
library(ggplot2)
library(dunn.test)

plot_dunn_ci <- function(df, pc = "PC1", grupo = "etapa", p_adjust = "bonferroni") {
  df[[grupo]] <- as.factor(df[[grupo]])
  
  dunn_res <- dunn.test(df[[pc]], g = df[[grupo]], method = p_adjust)
  
  pvals <- data.frame(
    Comparison = dunn_res$comparisons,
    P_adj = dunn_res$P.adjusted
  )
  
  # Medias y sd por grupo
  resumen <- df %>%
    group_by(.data[[grupo]]) %>%
    summarise(
      mean_pc = mean(.data[[pc]], na.rm = TRUE),
      sd_pc = sd(.data[[pc]], na.rm = TRUE),
      n = n(),
      .groups = "drop"
    )
  
  grupos <- levels(df[[grupo]])
  comparaciones <- t(combn(grupos, 2)) %>% as.data.frame()
  colnames(comparaciones) <- c("grupo1", "grupo2")
  
  # diferencia de medias e IC
  comparaciones <- comparaciones %>%
    rowwise() %>%
    mutate(
      mean1 = resumen$mean_pc[resumen[[grupo]] == grupo1],
      mean2 = resumen$mean_pc[resumen[[grupo]] == grupo2],
      sd1 = resumen$sd_pc[resumen[[grupo]] == grupo1],
      sd2 = resumen$sd_pc[resumen[[grupo]] == grupo2],
      n1 = resumen$n[resumen[[grupo]] == grupo1],
      n2 = resumen$n[resumen[[grupo]] == grupo2],
      Diff = mean1 - mean2,
      SE = sqrt(sd1^2 / n1 + sd2^2 / n2),
      Lower = Diff - 1.96 * SE,
      Upper = Diff + 1.96 * SE,
      Comparison = paste(grupo1, "-", grupo2)
    )
  
  # Unir p-valores de Dunn
  comparaciones <- comparaciones %>%
    left_join(pvals, by = "Comparison") %>%
    mutate(Signif = ifelse(P_adj <= 0.05, "*", ""))
  
  # Gráfico
  g <- ggplot(comparaciones, aes(x = Diff, y = Comparison)) +
    geom_point(size = 3) +
    geom_errorbarh(aes(xmin = Lower, xmax = Upper), height = 0.2) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
    geom_text(aes(label = Signif, x = Upper + 0.1), size = 5) +
    theme_minimal() +
    xlab(paste("Diferencia de medias (IC 95%)", pc)) +
    ylab("Comparaciones") +
    ggtitle(paste("Comparaciones post-hoc de", pc, "por", grupo)) +
    theme(plot.title = element_text(hjust = 0.5))
  
  return(list(
    dunn_results = dunn_res,
    summary_table = comparaciones,
    plot = g
  ))
}
```
```{r}
res_pc1 <- plot_dunn_ci(df_seg, pc = "PC1")
res_pc1$plot             
res_pc1$summary_table    
res_pc1$dunn_results    

```

```{r}
res_pc2 <- plot_dunn_ci(df_seg, pc = "PC2")
res_pc2$plot             
res_pc2$summary_table    
res_pc2$dunn_results    

```

# regresion logistica

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)

vars <- c("Factor_1","Factor_2","Factor_3")

df <- df %>%
  mutate(prob_hat = predict(m, type = "response"),
         logit_hat = qlogis(pmin(pmax(prob_hat, 1e-6), 1-1e-6)))   # avoid 0/1

cal_mod <- glm(avenida_cerrada ~ logit_hat, family = binomial, data = df)

calibrate_prob <- function(p, mod) {
  lp <- qlogis(pmin(pmax(p, 1e-6), 1-1e-6))
  plogis(coef(mod)[1] + coef(mod)[2] * lp)
}

make_curve <- function(var, n = 200) {
  others <- setdiff(vars, var)
  base   <- df %>% summarize(across(all_of(others), ~mean(.x, na.rm = TRUE)))
  rng    <- range(df[[var]], na.rm = TRUE)
  x      <- seq(rng[1], rng[2], length.out = n)

  grid <- base[rep(1, n), , drop = FALSE]
  grid[[var]] <- x

  p_raw <- predict(m, newdata = grid, type = "response")

  tibble(
    variable = var,
    x_raw = x,
    x_std = (x - mean(df[[var]], na.rm = TRUE)) / sd(df[[var]], na.rm = TRUE),
    p_raw = p_raw,
    p_cal = calibrate_prob(p_raw, cal_mod)
  )
}

curves <- map_dfr(vars, make_curve)

p_cal <- ggplot(curves, aes(x = x_std, y = p_cal, color = variable)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Predicted Probability vs. Factors",
    x = "Factor value",
    y = "Probaility",
    color = "Factor"
  ) +
  ylim(0, 1) +
  theme_minimal(base_size = 12)

print(p_cal)

ggsave("prob_curves_calibrated_centro.png", p_cal, width = 7, height = 5, dpi = 300)

```


